Supplementary materials

### 2018-08-13 WholePlotTrajectories.tex
mise en page

\JournalInfo{Hal xxx} % Journal information
\Archive{DOI xxxx} % Additional notes (e.g. copyright, DOI, review/research article)

@@

\Authors{
Ariane MIRABEL\textsuperscript{1*}\\ Eric Marcon\textsuperscript{1}\\ Bruno Hérault\textsuperscript{2 3}
} % Authors
\affiliation{
\textsuperscript{1}UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles,
Université de Guyane.\\ \hspace{1em} Campus Agronomique, 97310 Kourou, France.\\\textsuperscript{2}Cirad, Univ montpellier, UR Forests \& Societies.\\ \hspace{1em} Montpellier, France.\\\textsuperscript{3}INPHB, Institut National Polytechnique Félix Houphouet-Boigny\\ \hspace{1em} Yamoussoukro, Ivory Coast.
}
\affiliation{*\textbf{Corresponding author}: ariane.mirabel@ecofog.gf, http://www.ecofog.gf/spip.php?article47} % Corresponding author

\Keywords{Taxonomic and Functional Biodiversity, Neotropical Forests, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of communities'
resilience regarding their taxonomic and functional facets. To
disentangle the ecological processes driving community response to
disturbance, we analysed the tree community trajectories over 30 years
following a disturbance gradient in a Neotropical forest. Specifically,
we examined community functional and taxonomic trajectories with regards
to diversity, composition and redundancy. Functional trajectories were
drawn based on 7 leaf, stem and life-history traits. We highlighted the
cyclic recovery of community taxonomic and functional composition. While
pre-disturbance taxonomic differences were maintained over time,
functional composition trajectories were quite similar among
communities. The IDH did predict communities taxonomic diversity
response while functional diversity was enhanced whatever the
disturbance intensity. Although consistent, the recovery of community
composition, diversity and redundancy remained unachieved after 30
years. This acknowledged the need of decades-long cycles with no
disturbance to ensure a complete recovery, and questioned tropical
forest community resilience after repeated disturbances.
}

@@

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000}. One of the cornerstone of tropical forest ecology
is to understand the processes and drivers of ecosystems response to
disturbance \citep{Chazdon2003a}. For now, this has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016} that are
rapid and convenient to measure. These structural parameters have been
sucessfully modeled, giving important insights into the recovery of
ecosystem processes and services \citep{Herault2018}. However the
response of forests diversity and composition remains unclear, albeit it
determines the productivity, stability and functioning of ecosystems
\citep{Tilman2014, Liang2016}. In the short-term, moderate disturbance
may lead to positive impacts on communities diversity, an idea
formalized by the intermediate disturbance hypothesis (IDH) stating a
maximized species diversity when disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, however, gives limited or misleading information on
forests recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness,
that would reveal the changes in the species abundance distribution and
thus the underlying ecological processes, and (ii) composition that is
crucial for conservation issues \citep{Lavorel2002, Bellwood2006}.
Furthermore, a functional approach accounting for species biological
attributes would directly link communities diversity, composition and
redundancy to ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
sucessfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of communities response to disturbance would
encompass the changes in functional redundancy, that quantifies the
amount of shared trait values among species \citep{Carmona2016}. The
high functional redundancy of hyperdiverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) questioned
community recovery time.

@@

See Supplementary Materials -Figure S1 and \citet{Aubry-Kientz2013} for
the detailed methodology.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS using
abundance-based (Bray-Curtis) dissimilarity measures were conducted to
map either taxonomic or functional composition, the later based on the 7
leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the euclidean distance
between the target inventories and the reference inventories in 1989,
\emph{i.e.} 2 years after disturbance. Univariate trajectories of the
leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recomended for diversity studies \citep{Marcon2015b}. The
functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analysed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan with a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community. Community functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (Supp. Mat. - Figure S1 for a more
comprehensive sheme).

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
marked and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.87\) and
\(\rho_{spearman}^{functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

\caption{Plot trajectories in terms of flora composition (left panels \textbf{(a)} and \textbf{(c)}) and functional composition (right panels \textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the euclidean distance to initial condition along the 30 sampled years. Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

Maximum height at adult stage (\emph{Hmax}), leaf toughness
(\emph{L\_toughness}) and wood specific gravity (\emph{WSG}) first
decreased and then slightly increased but remained significantly lower
than their initial value (Figure \ref{fig:CWM}). On the other side, Bark
thickness (\emph{Bark\_thick}) and specific leaf area (\emph{SLA})
increased and while \emph{Bark\_thick} remained substantially high after
30 years, \emph{SLA} had almost recovered its initial value. For all
traits, the maximum difference to initial value was correlated to the
disturbance intensity (\(\rho_{spearman}^{L_{thickness}}=0.76\),
\(\rho_{spearman}^{L_{chloro}}=0.60\),
\(\rho_{spearman}^{L_{toughness}}=-0.53\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WSG}=-0.75\),
\(\rho_{spearman}^{Bark-thickness}=0.71\),
\(\rho_{spearman}^{Hmax}=-0.40\)). The proportions of the three lightest
seed mass classes increased in all disturbed plots, and decreased after
30 years for the lightest class while it stabilized for the two other
(Supp. Mat. - Figure S2).

@@

\caption{Trajectories of the communities weighted means (CWM) over 30 years after disturbance of 4 leaf traits (Leaf thickness, \emph{L\_thickness}, chlorophyll content, \emph{L\_chloro}, toughness, \emph{L\_toughness} and specific area, \emph{SLA}), 2 stem traits (wood specific gravity, \emph{WSG}, and bark thickness, \emph{Bark-thick}) and one life history trait (Specific maximum height at adult stage, \emph{Hmax}). Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:CWM}
\end{figure*}

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Colors are treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
functional richness and evenness trajectories depended on the
disturbance intensity with their maximum positively correlated to \%AGB
loss \(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to a maximum reached after 20-25 years, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the inital values.

The second-degree polynomial regressions between (i) the \%AGB loss and
(ii) taxonomic and functional richness and evenness after 10, 20 and 30
years best predicted the hump-shaped curve of the disturbance impact
along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the taxonomic
evenness. For both functional richness and evenness the relationship was
almost linear. The regression model better predicted the functional
richness and evenness (\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Evenness}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Evenness}<0.43\) respectively)

@@

\caption{Relationship between the initial \%AGB loss and community taxonmic richness \textbf{(a)}, taxonomic evenness \textbf{(b)}, functional richness \textbf{(c)},and functional evenness \textbf{(d)} at 10, 20 and 30 years after disturbance}\label{fig:IDHplot}
\end{figure*}

@@

\caption{Trajectories of the functional redundancy within the initial functional space over 30 years after disturbance. Colors are disturbance treatments: green (control), yellow (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:RedFunRest}
\end{figure}

@@

Communities taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years.

The taxonomic differences among local communities, marked before
disturbance by the distinct starting points on the NMDS axis 2, were
maintained throughout recovery trajectories. More than commonly thought,
post-disturbance trajectories depended on community initial composition,
that partly determined the pool of recruited species and constrained the
trajectories towards the initial composition. The high resilience of
communities taxonomy revealed that species not belonging to the
pre-disturbance community were hardly recruited because of the
commonness of dispersal limitation among tropical tree species
\citep{Svenning2005}.

Conversely, disturbed communities followed functional trajectories that
are highly similar in terms of functional composition. As
pre-disturbance surviving trees mirror the initial community
\citep{Herault2018}, changes in functional composition relied upon the
recruitment of species or functional types that were infrequent or
absent before disturbance. Competitive pioneers became dominant in
filling the environmental niches of high availability of light, space
and nutrients vacated by the disturbance. The recruitment of pioneers
changed community functional composition in the same way for all
disturbance intensity towards more resource-acquisitive strategies,
moving community functional composition right along the first axis in
Figure \ref{fig:NMDSplans} \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter long-lived, more resistant and shade-tolerant species
excluded the first established pioneers and started the recovery of
pre-disturbance functional composition, moving similarly community
functional composition left along the first axis and upward along the
second axis in Figure \ref{fig:NMDSplans}.

@@

The IDH well predicted well the disturbance impact on community
taxonomic richness, enhanced until an intensity threshold (20-25\% AGB
loss), and to some extent on taxonomic evenness, somewhat decoupled from
the disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.) \citep{Guitet2018}.
Below the intensity threshold the size of the surviving community
maintained the pre-disturbance high taxonomic richness while the
recruitment of pioneers, infrequent or absent before disturbance,
increased both community taxonomic richness and evenness. Beyond the
intensity threshold, the disturbance decreased the taxonomic richness of
surviving trees which was not offset by the enrichment of pioneers, so
that the overall community taxonomic richness decreased according to the
disturbance intensity \citep{Molino2001}. For community taxonomic
evenness the disturbance impact was similar but slighter, as the
evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pionneers that
balanced the usual hyper-dominance of a few species in tropical forests
below the intensity threshold, thus increasing community overall
evenness up to the intensity threshold beyond which pioneers became in
turn highly dominant and decreased the overall evenness
\citep{Baraloto2012a}.

Conversely the IDH was disproved regarding the disturbance impact on
community functional richness and evenness. Irrespective of the
disturbance intensity the recruitment of pionneers, functionally highly
different from the composition of pre-disturbance community, increased
both community functional richness and eveness.

Along time, taxonomic richness trajectories of all disturbed communities
first dropped similarly, following the species loss due to disturbance,
and then displayed a species gain depending on the disturbance
intensity. Up to an intensity threshold, the species gain was all the
more significant that the disturbance intensity increased, with the
establishment of long-lived pioneers enhancing community taxonomic
richness and evenness in the long term. These long-lived pionneers,
functionally quite different from the functional composition, entailed
as well a progressive and long-lasting increase of the functional
richness and evenness \citep{Denslow1980, Molino2001}. Beyond an
intensity threshold, though, a few short-lived pioneers occupied the
vacated environmental space and prevented the establishment of other
species. These short-lived pioneers were functionally very different
from the pre-disturbance community and entailed a rapid and significant
increase of functional richness en evenness. Already after 10 years,
though, short-lived pioneers started to decline and the functional
richness and evenness decreased. Likely this decrease will be followed
by the establishment of long-lasting pioneers, and by the time they
recruit we expect the taxonomic and functional trajectories to catch up
with those observed after intermediate disturbance \citep{Walker2009}.

>>>>

\JournalInfo{\ }
\Archive{\ }

@@

\Authors{
Ariane MIRABEL\textsuperscript{1*}\\ Eric Marcon\textsuperscript{1}\\ Bruno Hérault\textsuperscript{2 3}
} % Authors
\affiliation{
\textsuperscript{1}UMR EcoFoG, AgroParistech, CNRS, Cirad, INRA, Université des Antilles,
Université de Guyane.\\ \hspace{1em} Campus Agronomique, 97310 Kourou, France.\\\textsuperscript{2}Cirad, Univ montpellier, UR Forests \& Societies.\\ \hspace{1em} Montpellier, France.\\\textsuperscript{3}INPHB, Institut National Polytechnique Félix Houphoüet-Boigny\\ \hspace{1em} Yamoussoukro, Ivory Coast.
}
\affiliation{*\textbf{Corresponding author}: ariane.mirabel@ecofog.gf, https://github.com/ArianeMirabel} % Corresponding author

\Keywords{Community Ecology, Determinants of Plant Community Diversity and Structure, Disturbance Trajectories, Intermediate Disturbance Hypothesis, Long-term Resilience, Neotropical Forests, Taxonomic and Functional Biodiversity} % Keywords - if you don't want any simply remove all the text between the curly brackets
\newcommand{\keywordname}{Keywords} % Defines the keywords heading name

@@

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity. This intermediate disturbance hypothesis (IDH), though,
remains debated and the controversy questions the extent of communities'
resilience regarding their taxonomic and functional facets. To
disentangle the ecological processes driving community response to
disturbance, we analyzed the tree community trajectories over 30 years
following a disturbance gradient in a Neotropical forest. Specifically,
we examined community functional and taxonomic trajectories with regards
to diversity, composition and redundancy. Functional trajectories were
drawn based on 7 leaf, stem and life-history traits. We highlighted the
cyclic recovery of community taxonomic and functional composition. While
pre-disturbance taxonomic differences were maintained over time,
functional composition trajectories were quite similar among
communities. The IDH did predict communities taxonomic diversity
response while functional diversity was enhanced whatever the
disturbance intensity. Although consistent, the recovery of community
composition, diversity and redundancy remained unachieved after 30
years. This acknowledged the need of decades-long cycles with no
disturbance to ensure a complete recovery, and questioned tropical
forest community resilience after repeated disturbances.
}

@@

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events changing both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000}. One of the cornerstone of tropical forest ecology
is to understand the processes and drivers of ecosystems response to
disturbance \citep{Chazdon2003a}. For now, this has been largely studied
through forest structural parameters such as aboveground biomass, tree
height or stem density \citep{Piponiot2016, Rutishauser2016} that are
rapid and convenient to measure. These structural parameters have been
successfully modeled, giving important insights into the recovery of
ecosystem processes and services \citep{Herault2018}. However the
response of forests diversity and composition remains unclear, albeit it
determines the productivity, stability and functioning of ecosystems
\citep{Tilman2014, Liang2016}. In the short-term, moderate disturbance
may lead to positive impacts on communities diversity, an idea
formalized by the intermediate disturbance hypothesis (IDH) stating a
maximized species diversity when disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness alone, however, gives limited or misleading information on
forests recovery and functioning \citep{Chaudhary2016}. More
ecological-meaningful analysis would couple richness with (i) evenness,
that would reveal the changes in the species abundance distribution and
thus the underlying ecological processes, and (ii) composition that is
crucial for conservation issues \citep{Lavorel2002, Bellwood2006}.
Furthermore, a functional approach accounting for species biological
attributes would directly link communities diversity, composition and
redundancy to ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Baraloto2012a}. In that respect, the functional
trait-based approach that focus on major traits related to species
ecology and mediate species performance in a given environment was
successfully adopted \citep{Diaz2005}. For instance, the functional
approach revealed in tropical rainforests the deterministic processes
entailing, after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{Reich2014, Herault2011}. This shift is
translated into the trajectories of key functional traits related to
resource acquisition (leaf and stem traits) and life-history traits
(seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of community response to disturbance would encompass
the changes in functional redundancy, that quantifies the amount of
shared trait values among species \citep{Carmona2016}. The high
functional redundancy of hyper-diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
Neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem above-ground biomass (AGB) loss. We made
use of a large functional traits database encompassing major leaf, stem
and life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy.
Specifically, we (i) elucidated community taxonomic and functional
recovery and the underlying ecological processes, (ii) clarified the
validity of the IDH in the long term for tropical forest and its
translation into different trajectories in time, and (iii) questioned
community recovery time.

@@

See Supplementary Materials -Fig. S1 and \citet{Aubry-Kientz2013} for
the detailed methodology.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS using
abundance-based (Bray-Curtis) dissimilarity measures were conducted to
map either taxonomic or functional composition, the later based on the 7
leaf, stem and life history traits (without seed mass classes).
Trajectories along time were reported through the Euclidean distance
between the target inventories and the reference inventories in 1989,
\emph{i.e.} 2 years after disturbance. Univariate trajectories of the
leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007}. Species seed mass
corresponded to 5 classes of increasing mass, seed mass trajectories
were therefore reported as the proportion of each class in the
inventories (Supplementary materials).

The taxonomic diversity was reported through species richness and
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These indices belong to the set of HCDT or generalized
entropy, respectively corresponding to the 0 and 2 order of diversity
(q), recommended for diversity studies \citep{Marcon2015b}. The
functional diversity was reported using the functional richness and
functional evenness, \emph{i.e} Rao index of quadratic entropy which
combines species abundance distribution and average pairwise
dissimilarity based on species functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB loss. They were besides
analyzed through polynomial regression between (i) taxonomic and
functional richness and evenness and (ii) the initial \%AGB loss at 10,
20 and 30 years after disturbance.

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan with a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community. Community functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (Supp. Mat. - Fig. S1 for a more
comprehensive scheme).

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Fig. \ref{fig:NMDSplans}), they followed marked
and consistent trajectories over post-\break disturbance time. In
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WSG to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Fig. \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{Taxonomic}=0.87\) and
\(\rho_{spearman}^{Functional}=0.90\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

\caption{Plot trajectories in terms of flora composition (left panels \textbf{(a)} and \textbf{(c)}) and functional composition (right panels \textbf{(b)} and \textbf{(d)}) in a two-dimensional NMDS space. Lower panels (\textbf{(c)} and \textbf{(d)}) represent the Euclidean distance to initial condition along the 30 sampled years. Shaded areas are the credibility intervals.}\label{fig:NMDSplans}
\end{figure*}

@@

Maximum height at adult stage (\emph{Hmax}), leaf toughness and wood
specific gravity (\emph{WSG}) first decreased and then slightly
increased but remained significantly lower than their initial value
(Fig. \ref{fig:CWM}). On the other side, bark thickness and specific
leaf area (\emph{SLA}) increased and while bark thickness remained
substantially high after 30 years, \emph{SLA} had almost recovered its
initial value. For all traits, the maximum difference to initial value
was correlated to the disturbance intensity
(\(\rho_{spearman}^{Leaf thickness}=0.76\),
\(\rho_{spearman}^{Chlorophyll content}=0.60\),
\(\rho_{spearman}^{Leaf toughness}=-0.53\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WSG}=-0.75\),
\(\rho_{spearman}^{Bark thickness}=0.71\),
\(\rho_{spearman}^{Hmax}=-0.40\)). The proportions of the three lightest
seed mass classes increased in all disturbed plots, and decreased after
30 years for the lightest class while it stabilized for the two other
(Supp. Mat. - Fig. S2).

@@

\caption{Trajectories of the communities weighted means over 30 years after disturbance of four leaf traits (Leaf thickness, chlorophyll content, toughness, and specific area), two stem traits (wood specific gravity, and bark thickness) and one life history trait (Specific maximum height at adult stage). }\label{fig:CWM}
\end{figure*}

@@

\caption{Trajectories over 30 years of the difference with the 1989 inventory (2 years after disturbance) of community taxonomic \textbf{(a)} richness, \textbf{(b)}, taxonomic evenness, \textbf{(c)} functional richness, and \textbf{(d)} functional evenness. Shaded areas are the credibility intervals }\label{fig:DivTaxo}
\end{figure*}

The plot 7 from treatment 1 displayed constantly outlying functional
richness and evenness and was removed from the graphical representation
for better readability. In undisturbed plots both functional richness
and evenness remained stable along the 30 years. In disturbed plots,
functional richness and evenness trajectories depended on the
disturbance intensity with their maximum positively correlated to \%AGB
loss \(\rho_{spearman}^{Richness}=0.76\) and
\(\rho_{spearman}^{Evenness}=0.60\). Functional richness and evenness
displayed for low disturbance intensity a low but long-lasting increase
up to a maximum reached after 20-25 years, and for high intensity, a
fast but short increase followed after 10 years by a slow decrease
towards the initial values.

The second-degree polynomial regressions between (i) the percentage AGB
loss and (ii) taxonomic and functional richness and evenness after 10,
20 and 30 years best predicted the hump-shaped curve of the disturbance
impact along the disturbance intensity gradient \ref{fig:IDHplot}. The
relationship between the disturbance impact and its intensity was more
markedly hump-shaped for the taxonomic richness than for the taxonomic
evenness. For both functional richness and evenness the relationship was
almost linear. The regression model better predicted the functional
richness and evenness (\(0.55<R^2_{Functional Richness}<0.72\), and
\(0.60<R^2_{Functional Evenness}<0.81\)) than the taxonomic richness and
evenness (\(0.21<R^2_{Taxonomic Richness}<0.4\), and
\(-0.15<R^2_{Taxonomic Evenness}<0.43\) respectively)

@@

\caption{Relationship between the initial \%AGB loss and community taxonomic richness \textbf{(a)}, taxonomic evenness \textbf{(b)}, functional richness \textbf{(c)},and functional evenness \textbf{(d)} at 10, 20 and 30 years after disturbance}\label{fig:IDHplot}
\end{figure*}

@@

\caption{Trajectories of the functional redundancy within the initial functional space over 30 years after disturbance. Shaded areas are the credibility intervals.}\label{fig:RedFunRest}
\end{figure}

@@

Community taxonomic and functional composition appeared resilient,
following similar hump-shaped trajectories starting to return towards
pre-disturbance composition after 30 years.

The taxonomic differences among local communities, marked before
disturbance by the distinct starting points on the NMDS axis 2, were
maintained throughout recovery trajectories. More than commonly thought,
post-disturbance trajectories depended on community initial composition,
that partly determined the pool of recruited species and constrained the
trajectories towards the initial composition. The high resilience of
community taxonomy, in the sense of the initial state recovery, revealed
that species not belonging to the pre-disturbance community were hardly
recruited because of the commonness of dispersal limitation among
tropical tree species \citep{Svenning2005}.

Conversely, disturbed communities followed functional trajectories that
are highly similar in terms of functional composition. As
pre-disturbance surviving trees mirror the initial community
\citep{Herault2018}, changes in functional composition relied upon the
recruitment of species or functional types that were infrequent or
absent before disturbance. Competitive pioneers became dominant in
filling the environmental niches of high availability of light, space
and nutrients vacated by the disturbance. The recruitment of pioneers
changed community functional composition in the same way for all
disturbance intensity towards more resource-acquisitive strategies,
moving community functional composition right along the first axis in
Fig. \ref{fig:NMDSplans} \citep{Westoby1998, Wright2004, Reich2014}.
Thereafter long-lived, more resistant and shade-tolerant species
excluded the first established pioneers and started the recovery of
pre-disturbance functional composition, moving similarly community
functional composition left along the first axis and upward along the
second axis in Fig. \ref{fig:NMDSplans}.

@@

The IDH well predicted well the disturbance impact on community
taxonomic richness, enhanced until an intensity threshold (20-25\% AGB
loss), and to some extent on taxonomic evenness, somewhat decoupled from
the disturbance intensity as already observed in the Guiana Shield
\citep{Baraloto2012a} and in Bornean tropical forests
\citep{Cannon1998}. The disturbance intensity determined the balance in
the community between pre-disturbance surviving trees and those
recruited afterward. The pool of true pioneer species specifically
recruited after disturbance is restricted in the Guiana Shield to a few
common genera (e.g.~Cecropia spp., Vismia spp.) \citep{Guitet2018}.
Below the intensity threshold the size of the surviving community
maintained the pre-disturbance high taxonomic richness while the
recruitment of pioneers, infrequent or absent before disturbance,
increased both community taxonomic richness and evenness. Beyond the
intensity threshold, the disturbance decreased the taxonomic richness of
surviving trees which was not offset by the enrichment of pioneers, so
that the overall community taxonomic richness decreased according to the
disturbance intensity \citep{Molino2001}. For community taxonomic
evenness the disturbance impact was similar but slighter, as the
evenness is less sensitive to the loss of rare species. Taxonomic
evenness rather represented the increasing dominance of pioneers that
balanced the usual hyper-dominance of a few species in tropical forests
below the intensity threshold, thus increasing community overall
evenness up to the intensity threshold beyond which pioneers became in
turn highly dominant and decreased the overall evenness
\citep{Baraloto2012a}.

Conversely the IDH was disproved regarding the disturbance impact on
community functional richness and evenness. Irrespective of the
disturbance intensity the recruitment of pioneers, functionally highly
different from the composition of pre-disturbance community, increased
both community functional richness and evenness.

Along time, taxonomic richness trajectories of all disturbed communities
first dropped similarly, following the species loss due to disturbance,
and then displayed a species gain depending on the disturbance
intensity. Up to an intensity threshold, the species gain was all the
more significant that the disturbance intensity increased, with the
establishment of long-lived pioneers enhancing community taxonomic
richness and evenness in the long term. These long-lived pioneers,
functionally quite different from the functional composition, entailed
as well a progressive and long-lasting increase of the functional
richness and evenness \citep{Denslow1980, Molino2001}. Beyond an
intensity threshold, though, a few short-lived pioneers occupied the
vacated environmental space and prevented the establishment of other
species. These short-lived pioneers were functionally very different
from the pre-disturbance community and entailed a rapid and significant
increase of functional richness and evenness. Already after 10 years,
though, short-lived pioneers started to decline and the functional
richness and evenness decreased. Likely this decrease will be followed
by the establishment of long-lasting pioneers, and by the time they
recruit we expect the taxonomic and functional trajectories to catch up
with those observed after intermediate disturbance \citep{Walker2009}.
